---
title: "Richmond Player Performance"
output: html_document
---

### Aim
AFL is a uniquely Australian sport that forms a significant part of the Victorian identity. Founded in 1885, the Richmond Football Club is one of the longest running football clubs in Australia. Entering the Victorian Football League in 1908, the team has had a total of 1193 players play for the club to date. In this analysis, I will be using data retrieved from [afltables.com] (https://afltables.com/afl/stats/teams/richmond.html) to assess the performance of past and present players in the Richmond Football Club to ultimately identify our greatest player of all time.


### Load libraries
```{r load libraries}
library(tidyverse)
library(naniar)
library(shiny)
library(plotly)
```

### Read in data
```{r load data}
# Player performance data from 1965 to 2024
player_stats <-read.csv("Player Stats 1965-2024.csv")

# List of player information since the club's inception
player_list <-read.csv("Player List.csv")
```

### Data Exploration and Preparation
```{r initial exploration}
# View data dimensions
cat("Number of rows:", nrow(player_stats), "\n")
cat("Number of columns:", ncol(player_stats), "\n")

#Is there missing data?
cat("\n", "Amount of missing data per column:", "\n")
colSums(is.na(player_stats))
```
Our Player Stats dataset contains data for 550 players across 24 attributes. Some missing data is to be expected (e.g. a team's full back would rarely kick a goal), however certain columns have a lot more than is reasonable. When reviewing the data, we find that some attributes only started being recorded during the 1998 and 1999 AFL seasons. Therefore, we will need to split our data into 'core' attributes, recorded since 1965, and 'modern' attributes, recorded since 1998/1999.

Additionally, some statistics are redundant and will be removed from our dataset. For example, while we care about the number of marks a player has taken, whether they were contested or uncontested is not. Similarly, the number of Brownlow votes a player accrues reflects the umpires' opinions which may not reflect pure performance.

Before starting our analysis, we will also address any missing data by replacing null values with zeros. As we are searching for the best Richmond player of all time, players with less than 100 games will also be dropped. This hurdle has been set to align with the father-son rule, an established indicator of a player's high contribution to their club.

```{r data cleaning}
# drop columns
player_stats <- subset(player_stats, select = -c(HO, BR, CP, UP, CM, MI, BO, GA))
# rename X1. column
names(player_stats)[which(names(player_stats) == "X1.")] <- "1P"
# replace NA with 0
player_stats[is.na(player_stats)] <- 0
# drop players with less than 100 games
player_stats <- filter(player_stats, player_stats$GM >=100)
```

I will create a new column 'ACC' that uses the goals and behinds data to calculate each player's accuracy.
```{r accuracy metric}
# create accuracy column
player_stats <- mutate(player_stats, ACC = GL / (GL + BH))
# move accuracy column next to GL column
player_stats <- relocate(player_stats, ACC, .after=GL)
# drop behinds column
player_stats <- select(player_stats, "Player", "GM", "KI", "MK", "HB", "DI", "GL", "ACC", "TK", "RB", "IF", "CL", "CG", "FF", "FA", "1P")
```

While there is some subjectivity, free kicks are usually awarded to players for being first to the ball or for strong tackling and given for undisciplined play. In this analysis, a free kick differential and ratio will both be considered.

```{r free kick columns}
# create differential column
player_stats <- mutate(player_stats, FD = FF - FA)
player_stats <- relocate(player_stats, FD, .after=FA)
# create ratio column
player_stats <- mutate(player_stats, FR = FF / FA)
player_stats <- relocate(player_stats, FR, .after=FD)
# drop frees for and against columns
player_stats <- select(player_stats, "Player", "GM", "KI", "MK", "HB", "DI", "GL", "ACC", "TK", "RB", "IF", "CL", "CG", "FD", "FR", "1P")
```

The Disposals attribute is a summation of the Kicks and Handballs columns and as such will not be used on its own. I will instead combine this metric with the Clangers attribute to assess each player's disposal efficiency.
```{r disposal efficiency}
# create disposal efficiency column
player_stats <- mutate(player_stats, DE = (DI - CG)/DI*100)
player_stats <- relocate(player_stats, DE, .after=DI)
player_stats$DE[player_stats$CG == 0] <- 0
# drop disposals and clangers columns
player_stats <- select(player_stats, "Player", "GM", "KI", "MK", "HB", "DE", "GL", "ACC", "TK", "RB", "IF", "CL", "FD", "FR", "1P")
```

When evaluating player performances, the use of total values alone will skew our findings in favour of those with a higher number of games played. While this is warranted with more games and higher stats being a strong indicator of a better player, I want to balance this by incorporating per game averages into the analysis. This additional dataset will not include the GM, DE, ACC and FD columns.

```{r per game metrics and totals tables}
# create per game dataframe
player_stats_per_game <- player_stats
player_stats_per_game[,3:15] <- player_stats_per_game[, 3:15] / player_stats_per_game$GM
player_stats_per_game$FR <- player_stats_per_game$FR * player_stats_per_game$GM
#drop GM, DE, ACC, FD columns
player_stats_per_game <- select(player_stats_per_game, "Player", "KI", "MK", "HB", "GL", "TK", "RB", "IF", "CL", "FR", "1P")
#drop FR column from totals
player_stats_totals <- select(player_stats, "Player", "GM", "KI", "MK", "HB", "DE", "GL", "ACC", "TK", "RB", "IF", "CL", "FD", "1P")
```

Now that the data has been cleaned and prepared, I can return to resolving the issue of additional statistics being included for players from the 1998 season onward. I will create two sets of tables, one featuring the core stats collected since 1965 and the other including the full range of metrics. The analysis will initially use the core statistics to compare players from across the years after which I will expand its scope to incorporate the additional performance attributes to see how they affect the rankings of the modern players.

```{r create core stats tables}
# Extract core player stats into its own data set
core_ps_totals <- select(player_stats_totals, Player, GM, KI, MK, HB, GL, ACC, FD)
core_ps_per_game <- select(player_stats_per_game, Player, KI, MK, HB, GL, FR)
```

A data dictionary is provided below:

```{r data dictionary}
Column_Name <- c("Player", "GM", "KI", "MK", "HB", "DE", "GL", "ACC", "TK", "RB", "IF", "CL", "FD", "FR", "1P")
Description <- c("Player Name", "Games", "Kicks", "Marks", "Handballs", "Disposal Efficiency", "Goals", "Accuracy", "Tackles", "Rebound 50s", "Inside 50s", "Clearances", "Free Differential", "Free Ratio", "1 Percenters")
dict <- data.frame(Column_Name, Description)
print.data.frame(dict, row.names = FALSE, max = NULL)
```

# EDIT FROM HERE
### Player Analysis

Initially just the core stats
Best player will be determined in the following way:
Rank players according to totals by awarding points according to distribution. calculate z-scores for each attribute and player. I'll then normalise the z-scores on a range of 0-100 for ease of readability / interpretation. These values will then be summed to get a final total for each player. This will be completed for both the totals data and the per game data.

GL, ACC - half points as they are primarily aimed at forwards, looking to keep balance between statistics specific to certain positions. Look at distribution of goals, only players with more than median goals can receive accuracy points.

Create weightings for each metric and sum the total and per game values for each player
Combine the total and per game figures based on a weighting to produce one final ranking for each player

Rank player performance based on totals
Note: GL and ACC columns both reward those playing forward positions and are therefore worth half points for balance. Additionally, points for accuracy can only be received by players who have kicked greater than the median number of goals. Median used due to skewed distribution.

All attributes are positively skewed. Games example below.
```{r density plot}
ggplot(core_ps_totals, aes(x = GM)) +
  geom_density(fill = "skyblue", color = "black") +
  labs(title = "Density Plot of Games Played", x = "Number of Games Played per Player", y = "Density")
```

```{r rank player performance - totals}
# calculate z-scores for each stat
stats <- c('GM', 'KI', 'MK', 'HB', 'GL', 'ACC', 'FD')
for (stat in stats) {
  z_col <- paste0(stat, "_z")
  core_ps_totals[[z_col]] <- scale(core_ps_totals[[stat]])
}

# normalise z-scores on a 0-100 scale
for (stat in stats) {
  points_col <- paste0(stat, "_points")
  z_col <- paste0(stat, "_z")
  core_ps_totals[[points_col]] <- 50 + (core_ps_totals[[z_col]] * 10)
}

# Removing points from ACC where GL is lower than median
core_ps_totals <- mutate(core_ps_totals, ACC_points = ifelse(GL < median(GL), 0, ACC_points))

# GL and ACC columns are worth half points
core_ps_totals$GL_points <- core_ps_totals$GL_points/2
core_ps_totals$ACC_points <- core_ps_totals$ACC_points/2

# sum the points for each player in a new 'summed_points' column
points_cols <- paste0(stats, "_points")
core_ps_totals$summed_totals <- rowSums(core_ps_totals[points_cols])
```

On raw totals, Kevin Bartlett, an absolute legend of the Richmond football club comes out on top. Playing the most games out of any other player sees him score highly across all stats but this doesn't diminish his performance with his number of kicks being almost double the next best player.


Rank player performance based on per game averages
```{r rank player performance - per game}
# calculate z-scores for each stat
stats <- c('KI', 'MK', 'HB', 'GL', 'FR')
for (stat in stats) {
  z_col <- paste0(stat, "_z")
  core_ps_per_game[[z_col]] <- scale(core_ps_per_game[[stat]])
}

# normalise z-scores on a 0-100 scale
for (stat in stats) {
  points_col <- paste0(stat, "_points")
  z_col <- paste0(stat, "_z")
  core_ps_per_game[[points_col]] <- 50 + (core_ps_per_game[[z_col]] * 10)
} 

# sum the points for each player in a new 'summed_points' column
points_cols <- paste0(stats, "_points")
core_ps_per_game$summed_averages <- rowSums(core_ps_per_game[points_cols])
```
When looking at per game averages, the top two are neck and neck with Brett Deledio, having a more well rounded game, and Matthew Richardson, scoring high in goals and marks. 


Calculate final player points 
```{r player points}
Ranking <- inner_join(select(core_ps_totals, Player, summed_totals),
                       select(core_ps_per_game, Player, summed_averages),
                       by = "Player")
Ranking$final_ranking <- Ranking$summed_totals * 0.6 + Ranking$summed_averages * 0.4

```

Kevin Bartlett comes out on top as Richmond's greatest player but this doesn't give the full picture. Only three of the top ten players are from the pre-1998 era. Let's rerun this analysis using the full data set to give us a clearer picture on how the modern players stack up.


### Full stats
give a better picture of a players performance covering more aspects of the game

```{r rank player performance - FULL TOTALS}
# calculate z-scores for each stat
stats <- c("GM", "KI", "MK", "HB", "DE", "GL", "ACC", "TK", "RB", "IF", "CL", "FD", "1P")
for (stat in stats) {
  z_col <- paste0(stat, "_z")
  player_stats_totals[[z_col]] <- scale(player_stats_totals[[stat]])
}

# normalise z-scores on a 0-100 scale
for (stat in stats) {
  points_col <- paste0(stat, "_points")
  z_col <- paste0(stat, "_z")
  player_stats_totals[[points_col]] <- 50 + (player_stats_totals[[z_col]] * 10)
}

# Removing points from ACC where GL is lower than median
gl_median <- median(player_stats_totals$GL, na.rm = TRUE)
player_stats_totals$ACC_points <- ifelse(player_stats_totals$GL < gl_median, 0, player_stats_totals$ACC_points)

# GL and ACC columns are worth half points
player_stats_totals$GL_points <- player_stats_totals$GL_points/2
player_stats_totals$ACC_points <- player_stats_totals$ACC_points/2

# sum the points for each player in a new 'summed_points' column
points_cols <- paste0(stats, "_points")
player_stats_totals$summed_totals <- rowSums(player_stats_totals[points_cols])
```

Looking at the totals for the full stats, Trent Cotchin is number one, Dustin has moved from 5 to 2 (clearances and inside 50s), Joel Bowden 12 to 3 ( Rebound 50s).

Rank player performance based on per game averages
```{r rank player performance - FULL AVERAGES}
# calculate z-scores for each stat
stats <- c("KI", "MK", "HB", "GL", "TK", "RB", "IF", "CL", "FR", "1P")
for (stat in stats) {
  z_col <- paste0(stat, "_z")
  player_stats_per_game[[z_col]] <- scale(player_stats_per_game[[stat]])
}

# normalise z-scores on a 0-100 scale
for (stat in stats) {
  points_col <- paste0(stat, "_points")
  z_col <- paste0(stat, "_z")
  player_stats_per_game[[points_col]] <- 50 + (player_stats_per_game[[z_col]] * 10)
} 

# sum the points for each player in a new 'summed_points' column
points_cols <- paste0(stats, "_points")
player_stats_per_game$summed_averages <- rowSums(player_stats_per_game[points_cols])
```

Brett Deledio on top. Surprising, never top 5? but consistently good across the board.

Calculate final player points 
```{r FULL player points}
Ranking <- inner_join(select(player_stats_totals, Player, summed_totals),
                       select(player_stats_per_game, Player, summed_averages),
                       by = "Player")
totals_weight = 0.6
averages_weight = 0.4
Ranking$final_ranking <- Ranking$summed_totals * totals_weight + Ranking$summed_averages * averages_weight

```

Trent Cotchin best modern Richmond player


#Visualisation

Visualising the full data set only. User will be able to select which attributes they want to include (to view just the core stats / what they deem important).
First create a combined dataset to be used for the visualisation

```{r prepare datasets for merging}
# totals data
# drop z-score columns
merge_totals <- player_stats_totals %>% select(-matches("_z"))
# update column names for merger
new_names <- gsub("_points$", "TP", names(merge_totals))
names(merge_totals) <- new_names
# drop summed totals
merge_totals <- merge_totals[, 1:(ncol(merge_totals) - 1)]

# averages data
# drop z-score columns
merge_averages <- player_stats_per_game %>% select(-matches("_z"))
# update column names for merger
new_names <- gsub("_points$", "AvgP", names(merge_averages))
names(merge_averages) <- new_names
# drop summed averages
merge_averages <- merge_averages[, 1:(ncol(merge_averages) - 1)]

# merge datasets
combined_ps <- inner_join(merge_totals, merge_averages, by="Player")

# Rename Totals/Averages columns
#Create a copy of the original names
col_names <- names(combined_ps)
# Process columns 2 to 14
col_names[2:14] <- gsub("\\.x$", "", col_names[2:14])  # Remove .x
col_names[2:14] <- paste0(col_names[2:14], "T")        # Add T to the end
# Process columns 28 to 37
col_names[28:37] <- gsub("\\.y$", "", col_names[28:37])  # Remove .x
col_names[28:37] <- paste0(col_names[28:37], "Avg")        # Add T to the end
# Assign the new names to the data frame
names(combined_ps) <- col_names

```

Visualisation will give a breakdown of the points assigned to each player. Requires a points value to be aggregated at the atttribute level.
```{r player points by attribute}
# combine attribute point values
combined_ps$CGM <- (combined_ps$GMTP*totals_weight)
combined_ps$CKI <- (combined_ps$KITP*totals_weight)+(combined_ps$KIAvgP*averages_weight)
combined_ps$CMK <- (combined_ps$MKTP*totals_weight)+(combined_ps$MKAvgP*averages_weight)
combined_ps$CHB <- (combined_ps$HBTP*totals_weight)+(combined_ps$HBAvgP*averages_weight)
combined_ps$CDE <- (combined_ps$DETP*totals_weight)
combined_ps$CGL <- (combined_ps$GLTP*totals_weight)+(combined_ps$GLAvgP*averages_weight)
combined_ps$CACC <- (combined_ps$ACCTP*totals_weight)
combined_ps$CTK <- (combined_ps$TKTP*totals_weight)+(combined_ps$TKAvgP*averages_weight)
combined_ps$CRB <- (combined_ps$RBTP*totals_weight)+(combined_ps$RBAvgP*averages_weight)
combined_ps$CIF <- (combined_ps$IFTP*totals_weight)+(combined_ps$IFAvgP*averages_weight)
combined_ps$CCL <- (combined_ps$CLTP*totals_weight)+(combined_ps$CLAvgP*averages_weight)
combined_ps$CFD <- (combined_ps$FDTP*totals_weight)
combined_ps$CFR <- (combined_ps$FRAvgP*averages_weight)
combined_ps$C1P <- (combined_ps$`1PTP`*totals_weight)+(combined_ps$`1PAvgP`*averages_weight)

# drop original total points columns
combined_ps <- combined_ps[, -c(15:27)]
# drop original average points columns
combined_ps <- combined_ps[, -c(25:34)]
#create ranking column
combined_ps$final_rankings <- rowSums(combined_ps[25:38])
```

Data can now be loaded into Shiny app

```{r interactive visualisation}
# Load my data
playerdata <- combined_ps

# Define UI
ui <- fluidPage(

  titlePanel("Analysing the performance of the Richmond Football Club"),
  sidebarLayout(
    sidebarPanel(
      # create user input for plot1
      checkboxGroupInput("values", "Select the Metrics to Include:",
                         choices = c("Games Played" = "CGM", 
                                     "Kicks" = "CKI", 
                                     "Marks" = "CMK", 
                                     "Handballs" = "CHB", 
                                     "Disposal Efficiency" = "CDE", 
                                     "Goals" = "CGL", 
                                     "Accuracy" = "CACC", 
                                     "Tackles" = "CTK", 
                                     "Rebound 50s" = "CRB", 
                                     "Inside 50s" = "CIF", 
                                     "Clearances" = "CCL",
                                     "Free Kick Differential" = "CFD",
                                     "Free Kick Ratio" = "CFR", 
                                     "1 Percenters" = "C1P"),
                        selected = c("CGM", "CKI", "CMK", "CHB", "CDE", "CGL", 
                                      "CACC", "CTK", "CRB", "CIF", "CCL", "CFD",
                                      "CFR", "C1P")),
      
    ),
  
    mainPanel(
      h4("Who is the greatest Richmond player of all time?"),
      p("By ranking players based on their per game and total metrics, we aim to
      identify Richmond's greatest player. Focus on the metrics that matter to 
      you most by selecting your own combinations using the checkboxes on the 
      left hand side."),
      plotlyOutput("barChart"),
      
    ))
  )

# Define server logic
server <- function(input, output) {
  
  # specify unique colors to avoid repeats
  c1 <- "#c7dbf0"
  c2 <- "#87a353"
  c3 <- "#f5c998"
  
  # plot1
  output$barChart <- renderPlotly({
    # rank players by totals
    sorted_data <- playerdata[order(playerdata$final_rankings), ]

    # create stacked bar chart
    performance <- plot_ly(sorted_data, y = ~Player, height = 700)
    
    # set each attribute within an if statement to allow reordering based on user selection
    if ("CGM" %in% input$values) {
      performance <- performance %>% add_trace(x = ~CGM, name = "Games", type = "bar", orientation = 'h',
                                               hoverinfo = "text", text = ~paste("Games: ", GMT),
                                               textposition = "none")
    }
    if ("CKI" %in% input$values) {
      performance <- performance %>% add_trace(x = ~CKI, name = "Kicks", type = "bar", orientation = 'h',
                                               hoverinfo = "text", text = ~paste("Kicks: ", KIT, "<br>", "Kicks per Game: ", KIAvg),
                                               textposition = "none")
    }
    if ("CMK" %in% input$values) {
      performance <- performance %>% add_trace(x = ~CMK, name = "Marks", type = "bar", orientation = 'h',
                                               hoverinfo = "text", text = ~paste("Marks: ", MKT, "<br>", "Marks per Game: ", MKAvg),
                                               textposition = "none")
      }
    if ("CHB" %in% input$values) {
        performance <- performance %>% add_trace(x = ~CHB, name = "Handballs", type = "bar", orientation = 'h',
                                                 hoverinfo = "text", text = ~paste("Handballs: ", HBT, "<br>", "Handballs per Game: ", HBAvg),
                                                 textposition = "none")
      }
      if ("CDE" %in% input$values) {
        performance <- performance %>% add_trace(x = ~CDE, name = "Disposals Efficiency", type = "bar", orientation = 'h',
                                                 hoverinfo = "text", text = ~paste("Disposals Efficiency: ", CDE),
                                                 textposition = "none")
      }
      if ("CGL" %in% input$values) {
        performance <- performance %>% add_trace(x = ~CGL, name = "Goals", type = "bar", orientation = 'h',
                                                 hoverinfo = "text", text = ~paste("Goals: ", GLT, "<br>", "Goals per Game: ", GLAvg),
                                                 textposition = "none")
      }
      if ("CACC" %in% input$values) {
        performance <- performance %>% add_trace(x = ~CACC, name = "Accuracy", type = "bar", orientation = 'h',
                                                 hoverinfo = "text", text = ~paste("Accuracy: ", ACCT),
                                                 textposition = "none")
      }
      if ("CTK" %in% input$values) {
        performance <- performance %>% add_trace(x = ~CTK, name = "Tackles", type = "bar", orientation = 'h',
                                                 hoverinfo = "text", text = ~paste("Tackles: ", TKT, "<br>", "Tackles per Game: ", TKAvg),
                                                 textposition = "none")
      }
      if ("CRB" %in% input$values) {
        performance <- performance %>% add_trace(x = ~CRB, name = "Rebound 50s", type = "bar", orientation = 'h',
                                                 hoverinfo = "text", text = ~paste("Rebound 50s: ", RBT, "<br>", "Rebound 50s per Game: ", RBAvg),
                                                 textposition = "none")
      }
      if ("CIF" %in% input$values) {
        performance <- performance %>% add_trace(x = ~CIF, name = "Inside 50s", type = "bar", orientation = 'h',
                                                 hoverinfo = "text", text = ~paste("Inside 50s: ", IFT, "<br>", "Inside 50s per Game: ", IFAvg),
                                                 textposition = "none")
      }
          if ("CCL" %in% input$values) {
        performance <- performance %>% add_trace(x = ~CCL, name = "Clearances", type = "bar", orientation = 'h',
                                                 hoverinfo = "text", text = ~paste("Clearances: ", CLT, "<br>", "Clearances per Game: ", CLAvg),
                                                 textposition = "none")
      }
      if ("CFR" %in% input$values) {
        performance <- performance %>% add_trace(x = ~CFR, name = "Frees", type = "bar", orientation = 'h',
                                                 hoverinfo = "text", text = ~paste("Ratio of free kicks given vs awarded: ", FRAvg),
                                                 textposition = "none", color=I(c1))
      }
      if ("CFD" %in% input$values) {
        performance <- performance %>% add_trace(x = ~CFD, name = "Free Differential", type = "bar", orientation = 'h',
                                                 hoverinfo = "text", text = ~paste("Free Kick Differential: ", FDT),
                                                 textposition = "none", color=I(c2))
      }
      if ("C1P" %in% input$values) {
        performance <- performance %>% add_trace(x = ~C1P, name = "1%", type = "bar", orientation = 'h',
                                                 hoverinfo = "text", text = ~paste("1%: ", `1PT`, "<br>", "1% per Game: ", `1PAvg`),
                                                 textposition = "none", color=I(c3))
      }
      
    performance <- performance %>% layout(barmode = 'stack', yaxis = list(title = "Players", categoryorder = "total ascending"), xaxis = list(title = "Rankings"))

  })
   
   
}

# Run the application
shinyApp(ui = ui, server = server)

```

